Revealing transient strain in geodetic data with Gaussian process regression

### 2017-09-28 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed stochastic process that is fully defined in terms of a mean function and a positive definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to objectively choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t$. We describe this observation as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d^{(j)}_i(t) = &u_i(\vec{x}^{(j)},t) + \eta_i(\vec{x}^{(j)},t) + a_i^{(j)} + b_i^{(j)}t + c_i^{(j)}\sin(2 \pi t) + \\ 
               &e_i^{(j)}\cos(2 \pi t) + f_i^{(j)}\sin(4 \pi t)  + g_i^{(j)}\cos(4 \pi t), 
\end{split}
\end{align}
where $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $\eta_i$ to represent all the stochastic noise processes. For example, $\eta_i$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that each component of $\eta_i$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

@@

We can use eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to evaluate $\hat{u}_i$ at any $p$. Furthermore, since $\hat{u}_i$ is spatially and temporally differentiable, we can use $\hat{u}_i$ to formulate $\dot\varepsilon$, which can also be evaluated at any $p$. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $u_i$ and $u_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
\end{equation} 
respectively.

Our motivation for estimating transient strain rates is, in part, to detect geophysical phenomena. As we will see, geophysical signal can be easily identified by visually inspecting the solution for transient strain rates. However, if we want to detect geophysical phenomena automatically, then we need to define a detection criterion. We propose using a signal-to-noise ratio, SNR, based on $\dot\varepsilon$ for our detection criterion. The Frobenius norm of $\dot\varepsilon$, $||\dot\varepsilon||_\mathrm{F} = \left(\dot\varepsilon_{\mathrm{ee}}^2 + \dot\varepsilon_{\mathrm{nn}}^2 + 2\dot\varepsilon_{\mathrm{en}}^2\right)^{\frac{1}{2}}$, which is sometimes referred to as the second invariant of strain rate in the geodetic literature, is often used as a metric for the strain rate ``magnitude''. Noting that $||\dot\varepsilon||_\mathrm{F}$ is a random variable, SNR can be taken as the ratio of the estimated mean and standard deviation of $||\dot\varepsilon||_\mathrm{F}$. Using nonlinear uncertainty propagation, we find SNR to be  
\begin{equation}\label{eq:SNR}
\mathrm{SNR}(p) = \frac{\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 +
                        \mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 +
                        2\mu_{\dot\varepsilon_\mathrm{en}}(p)^2}
                       {\big(C_{\dot\varepsilon_\mathrm{nn}}(p,p)\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 + 
                              C_{\dot\varepsilon_\mathrm{ee}}(p,p)\mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 + 
                              4C_{\dot\varepsilon_\mathrm{en}}(p,p)\mu_{\dot\varepsilon_\mathrm{en}}(p)^2
                        \big)^{\frac{1}{2}}}
,
\end{equation}
For simplicity, we have ignored covariances between the transient strain rate components in eq. (\ref{eq:SNR}), even though they are non-zero.

>>>>

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed stochastic process that is fully defined in terms of a mean function and a positive definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to objectively choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t$. We describe this observation as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d^{(j)}_i(t) = &u_i(\vec{x}^{(j)},t) + \eta_i(\vec{x}^{(j)},t) + a_i^{(j)} + b_i^{(j)}t + c_i^{(j)}\sin(2 \pi t) + \\ 
               &e_i^{(j)}\cos(2 \pi t) + f_i^{(j)}\sin(4 \pi t)  + g_i^{(j)}\cos(4 \pi t), 
\end{split}
\end{align}
where $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $\eta_i$ to represent all the stochastic noise processes. For example, $\eta_i$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that $\eta_i$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

@@

The posterior transient displacements are spatially and temporally continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to evaluate $\hat{u}_i$ at any $p$. Furthermore, $\hat{u}_i$ is spatially and temporally differentiable, allowing us to formulate $\dot\varepsilon$ at any $p$ that we may be interested in. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $u_i$ and $u_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
\end{equation} 
respectively.

Our motivation for estimating transient strain rates is, in part, to detect geophysical phenomena. As we will see, geophysical signal can be easily identified by visually inspecting the solution for transient strain rates. However, if we want to detect geophysical phenomena automatically, then we need to define a detection criterion. We use a signal-to-noise ratio, SNR, that is based on the Frobenius norm of $\dot\varepsilon$, $||\dot\varepsilon||_\mathrm{F} = \left(\dot\varepsilon_{\mathrm{ee}}^2 + \dot\varepsilon_{\mathrm{nn}}^2 + 2\dot\varepsilon_{\mathrm{en}}^2\right)^{\frac{1}{2}}$, for our detection criterion. In the geodetic literature, $||\dot{\varepsilon}||_\mathrm{F}$ is often used as a metric for the strain rate ``magnitude", and it is sometimes referred to as the second invariant of strain rate. Noting that $||\dot\varepsilon||_\mathrm{F}$ is a random variable, we take SNR to be the ratio of the estimated mean and standard deviation of $||\dot\varepsilon||_\mathrm{F}$. An estimate of the mean is found by evaluating $||\dot\varepsilon||_\mathrm{F}$ at the mean of $\dot\varepsilon$, 
\begin{align}
\mu_{||\dot\varepsilon||_\mathrm{F}} &\approx \left. ||\dot\varepsilon||_F \right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \\
                                     &= \left(\mu_{\dot\varepsilon_\mathrm{nn}}^2 + 
                                              \mu_{\dot\varepsilon_\mathrm{ee}}^2 + 
                                              2\mu_{\dot\varepsilon_\mathrm{en}}^2
                                     \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard deviation of $||\dot\varepsilon||_\mathrm{F}$,
\begin{equation}
\sigma_{||\dot\varepsilon||_\mathrm{F}} \approx
\left(\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{nn}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\varepsilon_\mathrm{nn}} +
\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{ee}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\varepsilon_\mathrm{ee}} +
\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{en}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\varepsilon_\mathrm{en}}
\right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\varepsilon_{ij}}(p) = C_{\varepsilon_{ij}}(p,p)$. 
 by evaluating it at $\dot\varepsilon_{ij} = \mu_{\dot\varepsilon_{ij}}$. point $p$ is approximately $(\mu_{\dot\varepsilon_\mathrm{nn}}(p) + \mu_{\dot\varepsilon_\mathrm{ee}}(p) + 2\mu_{\dot\varepsilon_\mathrm{en}}(p))$. Using nonlinear uncertainty propagation, the standard deviation of $||\dot\varepsilon||$ is approximated as

can be found with nonlinear uncertainty propagation,

 Using nonlinear uncertainty propagation, we find SNR to be  
\begin{equation}\label{eq:SNR}
\mathrm{SNR}(p) = \frac{\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 +
                        \mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 +
                        2\mu_{\dot\varepsilon_\mathrm{en}}(p)^2}
                       {\big(C_{\dot\varepsilon_\mathrm{nn}}(p,p)\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 + 
                              C_{\dot\varepsilon_\mathrm{ee}}(p,p)\mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 + 
                              4C_{\dot\varepsilon_\mathrm{en}}(p,p)\mu_{\dot\varepsilon_\mathrm{en}}(p)^2
                        \big)^{\frac{1}{2}}}
,
\end{equation}
For simplicity, we have ignored covariances between the transient strain rate components in eq. (\ref{eq:SNR}), even though they are non-zero.
